---
title: "Correlation in R"
output: pdf_document
---

## Goals
1. Illustrate running (Pearson) correlation in R
2. Illustrate running Spearman rank correlation (a non-parametric version of correlation) in R


Let's illustrate the correlation coefficient.
'sd_y' is the standard deviation of the y variable.
'a' allows us to make x and y correlated, or not.
```{r}
a =-1
sd_y = 3
x<-rnorm(30,mean=0,sd=3)
y<- rnorm(30,mean=0,sd=sd_y) + a*x

plot(x,y)
cor_xy = round(cor(x,y,method="pearson"),3)
text(0,0, cor_xy,col="red",cex = 2)
```



# A simple example of Pearson correlation. 
Let's create a very small data set with x and y variables.

```{r}
x <- c(1, 3, 4, 5, 7)
y <- c(5, 9, 7, 1, 13)
```

The `cor` command 
```{r}
cor(x,y,method="pearson")
```


```{r}
plot(x,y)
```

This type of correlation is also called a Pearson correlation. 

```{r}
cor(x,y,method="pearson")
```
The 'cor.test' command will test the null hypothesis that the correlation is equal to zero. 
```{r}
cor.test(x,y,method="pearson") 
```

The non-parametric version of the correlation test is the spearman rank correlation.  It is based on the ranks of the data, rather than the data values. 

```{r}
rankx <- rank(x)
rankx
ranky <- rank(y)
ranky
```

Performing a Pearson correlation test on the ranks is equivalent to running a Spearman correlatin test. 

```{r}
cor(rankx,ranky,method="pearson")
```


```{r}
cor(x,y,method="spearman")
```

