---
title: "Lab 4 video tutorial - R tricks & ttests"
author: "sbsambado"
date: "4/17/2020"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
#knit:: opts_chunks$set(root.dir = "~/Desktop/Classes/EEMB 146 S20/EEMB146_Biometry")

# watch Tatums awesome video of how to set your working directory!
```

The objectives of this video
1. How to turn .txt file into .Rmd file
2. Helpful R tricks
3. T-tests in R


## Helpful R tricks
1. Resources
  - Q&A forum on piazza
  - weblinks posted on GS and piazza
  - google really is your best friend!
~ you don't have to know everything, just know how to troubleshoot errors!

 
2. How to begin a code if you have no idea where to begin 
  *Remember* 
    variable_name <- function(argument(s))
    
        a. dataset: mammals
        b. variable: no.species
        c. specific variable: vertebrates
        d. function: sum 
        e. argument: " == 'yes' " 

  *Helpful hints*
      - ` ?function()`
```{r}
?t.test()
?t.test()



```

      - `$` refers to element in column/list within a data frame
      
      - " quotations " for character data (should make green writting)
      
      - `c()` concatenate (i.e. bring together multiple items)
      
      - `[ ] ` selecting data 
      
      - `+` you need to complete a code
      
      - `>` your code ran successfully!


3. Typical coding errors
   - spelling (readr != reader)
   - capitilization  ( readr != Readr)
   - punctuation (read.csv != read_csv)
   - open brackets
   - View() # run View() in consule or # View() after running code
   - `read.csv` not `read_csv` (it's a warning, but easier to fix it) 
   ~ unfortunately, you will not be able to knit a document if there's an error
        - however, the error message usually guides you to where the error is!
   
4. Good practice
  - # annotate your code 
  - Save your exercises for personal references, esp. for individual final projects

## T-test in R

`t.tests()` function produces a variety of t-tests

1. Types of t-test
  a. One sample t-test
  b. Paired t-test
  c. Two-sample t-test
  
  *One sample t-test*
    - Try to find the mean of a sample (without knowing variance of population)
        Assumes:
          - Random variable has normal distribution
          - The observations represent a random sample from the population
```{r one sample t-test}
## the one-sample t-test

# base code
# Ho: mu = 3
t.test(y, mu = 3)

# modify the code for your variables!
t.test(dataset$sample, mu = mu)
```

  *Paired t-test*
    - Compare means between two groups (each data point is uniquely paired with another group)
        Assumes:
          - The observations from each group represent a random sample from population
          - The difference of the two observations follow a normal distribution
```{r paired t-test}
## paired t-test

# base code
t.test(y1, y2, paired = TRUE) # where y1 and y2 are numeric
```
    
  *Two-sample t-test*
    - Compares whether the means of two groups are signficantly different from each other
      Assumes:
        - Observation from each group represents a random sample from the population
        - Observations follow a normal distribution
        - Observations from the two groups have the same variance
```{r two-sample t-test}
## independent 2-group t-test

# base code, 1 numeric value
t.test(y~x) # where y is numeric and x is a binary factor

# base code, 2 numeric values
t.test(y1, y2) # where y1 and y2 are numeric
```

2. Resource
  - 'ComparingMeansFlowchart.pptx.pdf' on GS


