---
title: "Lab 2 - Assignment"
output:
  html_document:
    df_print: paged
---

Lab 2: Introduction to R Markdown and Data Analysis

# 3 common syntax errors
# spelling 
install.packages('readr'), correct
instal.package('redr') , not correct

# capitilization
install.packages('redr'), correct
install.packages('redr'), not correct

# punctuation
install.packages('redr'), correct
instll_packages('readr'), not correct


Exercise 1: Bat tongue Data
```{r}
#install.packages('readr') # if first time using readr

## hashtag is making a comment, it is not seen as code
# annotating your code is great practice! 

caffeine <- read.csv("~/Desktop/caffeine.csv")

# upload data from gauchospace
library(readr)
bat_tongues <- read.csv("~/Desktop/Classes/EEMB 146 S20/Week 2/Lab 2/Lab 2 Files/battongues.csv")

# View(bat_tongues) don't include View() or it will not knit!

# make a copy of bat_tongues dataframe
bat_copy <-bat_tongues
# bat copy is dataframe that we're working with

# change names of columns in bat_copy
names(bat_copy) = c('species', 'palate', 'tongue')
#View(bat_copy)

# calculate means of 'tongue'
mean(bat_copy$tongue)

# summary statistics for entire data frame or variable
summary(bat_copy$tongue)
summary(bat_copy$species)

```

Exercise 2: Caffeine Data
```{r}
# transformed xcel file to csv

caffeine <- read.csv("~/Desktop/Classes/EEMB 146 S20/Week 2/Lab 2/Lab 2 Files/caffeine.csv")

# basic statistics
summary(caffeine)

# variance
var(caffeine$Caffeine..mg..in.16.oz..coffee)

# try code that Dr. Briggs went over in lecture during Week 1 (Wednesday)
mean(caffeine$Caffeine..mg..in.16.oz..coffee)
max(caffeine$Caffeine..mg..in.16.oz..coffee)
min(caffeine$Caffeine..mg..in.16.oz..coffee)
sd(caffeine$Caffeine..mg..in.16.oz..coffee)
median(caffeine$Caffeine..mg..in.16.oz..coffee)
range(caffeine$Caffeine..mg..in.16.oz..coffee)
quantile(caffeine$Caffeine..mg..in.16.oz..coffee)

# categorical data to analyze 
table(caffeine)
table(caffeine$Brand)

# plot 
coffee <-hist(caffeine$Caffeine..mg..in.16.oz..coffee, xlab = 'brand', ylab = 'caffeine',
              main = 'coffee',col = 'blue') 

```

## Homework to turn in 

Question 1. Mammal Data
```{r}


# upload data
mammals <- read.csv("~/Desktop/Classes/EEMB 146 S20/Week 2/Lab 2/Lab 2 Files/mammals.csv")
View(mammals)
# a. Which of the mammals has the largest brain size:body ratio?
brain_body <- mammals$brain.mass..g./mammals$body.mass..kg.

# combine columns
new_mammals<-cbind(mammals, brain_body)

# find max brain/body ratio
new_mammals[which.max(new_mammals$brain_body),]

# find min brain/body ratio
new_mammals[which.min(new_mammals$brain_body),]

# mean of brain/body ratio
mean(new_mammals$brain_body)

# find humans mean
# make new dataframe to select human
human <- new_mammals[which(new_mammals$name == 'Human'),] 
mean(human$brain_body)

# mammal mean is 9.5, Human mean is 21.3


## Alternative way ##

# rename data

mammals<- read.csv()
names(mammals) = c('name', 'body', 'brain')

# calculate new ratio variable
mammals['ratio'] = mammals$brain/ mammals$body

# Which mammal has the largest brain size: body size ratio?
largest_brain <- mammals[which.min(mammals$ratio),]

# Find the men brain size: body size ratio. Are humans above or below the mean?
mean(mammals$ratio) # finding the mean

  
mean_ratio <- mean(mammals$ratio)
print(mean_ratio)

print(5)


mammals[mammals$name == 'Human', 'ratio'] # finding humans ratio

# Making a logical statement
mean(mammals$ratio) < mammals[mammals$name == 'Human', 'ratio']
```

Question 2: Bird data
```{r}
desert_birds <- read.csv("~/Desktop/Classes/EEMB 146 S20/Week 2/Lab 2/Lab 2 Files/desert_birds.csv")

### Question 2.a how many rare (<10 sightings) birds are there? ###
rare <- subset(desert_birds, Count <= 10)
print(rare)
### ANSWER 2.a: There are 16 rare birds

# summary statistics
length(rare) # columns
dim(rare) # rows, columns
mean(rare$Count)
min(rare$Count)
max(rare$Count)


## Alternative Way ##

sum(birds$Count <= 10)

nrow(rare)
dim(rare)

```

Question 3: Endangered species data
```{r}
endangered <- read.csv("~/Desktop/Classes/EEMB 146 S20/Week 2/Lab 2/Lab 2 Files/endangered.csv")

# What proportion of these endangered species are vertebrates using R?

# make new data frame with just vertebrate
vertebrate <- subset(endangered, vertebrate == 'yes')
print(vertebrate)
View(vertebrate)

# find the sum of vertebrates
vertebrate_count <-sum(vertebrate$no.species)

# another way to 'count' your data
install.packages('plyr') # new package alert!
library(plyr) # always call back in that new package
count(vertebrate$no.species) # I am applying the function 'count' to data set = 

# Find the proportion of vertebrates that considered endangered 
endangered_count <- sum(endangered$no.species)

proportion <- vertebrate_count/endangered_count
print(proportion)

# vertebrates make up 0.27 or 27% of the endangered species list


## Alternative way

verts <- sum(endangered$no.species[endangered$vertebrate == 'yes'])
prop <- verts/sum(endangered$no.species)
prop
```

Quick Note based on Week 2 Lectures

## Distributions in R
In R, every distribution has four functions

1) Root name: for normal distribution this is norm
2) Prefix:
- p (for 'probability'): cumulative distribution function ,cdf pnorm
- q (for 'quantile'): inverse cdf qnorm
- d (for 'density'): the density function, pdf dnorm
- f (for 'random'): a random number from that distribution rnorm

Density function
```{r}
y = seq(-5, 5, 0.1)
fy = dnorm(y, mean = 0, sd = 1)
plot(y, fy, type = 'l', ylab = 'f(Y)', xlab = 'Y',
     main = 'Normal distribution')
```

Cumulative distribution function
```{r}
y = seq(-5, 5, 0.1)
cdfy = pnorm(y, mean = 0, sd = 1)
plot(y, cdfy, type = 'l', ylab = 'f(Y)', xlab = 'cumulative distribution',
     main = 'Normal distribution')
```

Quantiles
```{r}
# at what value of Y is the 50% quantile?
qnorm(0.5, mean = 0, sd = 1)

# At what value of Y is the 70% quantile?
x <- qnorm(0.7, mean = 0, sd = 1)

plot(y, cdfy, type = 'l', ylab = 'f(Y)', xlab = 'cumulative distribution',
     main = 'Normal distribution')
abline(v = x)
```

Random draws
```{r}
# give me 10 random draws from normal distribution

rnorm(10, mean = 0, sd = 1)
```

Log normal distribution
- continuous
- range: real values > 0 
- parameters: mu (mean of logarithm), sigma (sd of logarithm)
```{r}

y = seq(0,10, 0.1)
fy = dlnorm(y, mean = 0, sd = 1)
plot(y, fy, type = 'l', xlab = 'Y', main = 'log normal distribution')
```

t-distribution
```{r}
x = seq(-5, 5, 0.1)
y = dnorm(x, mean = 0, sd = 1)
plot(x, y, type = 'l', col = 'red',xlab = 'Y', ylab = 'f(Y)', main = 't distribution')
y2 = dt(x, df = 2)
lines(x, y2, col = 'blue')
```

Uniform distribution
- continuous
- range: a to b
- parameters: a (minimum), b (maximum)
```{r}
y = seq(0,10, 0.1)
fy = dunif(y, min = 2, max = 8)
plot(y, fy, type = 'l', xlab = 'Y', ylab = 'f(Y)', main = 'uniform')
```

Gamma distribution
- waiting time until a certain number of evetns take place
- range: positive real values
- parameters: scale (length per event [ rate = rate at which events occur ~ 1/scale]), shape (number of events)

e.g. gamma(shape = 3, scale = 2) is the distribution of time (in days) you'd expect to have to wait for 3 deaths in a population, given the average survival time is 2 days
```{r}
y = seq(0, 10 , 0.1)
fy = dgamma(y, shape = 3, scale = 2)
plot(y, fy, type = 'l', ylab = 'f(Y)', xlab = 'Y', main = 'gamma(shape = 3, scale = 2)')
```

Binomial distribution (0/1)
- number of successes out of a given number of trials
- discrete
- range: non-negative integers
- parameters: prob (prob. of each success), size (number of trials)
```{r}
y = seq(0,10)
fy = dbinom(y, 10, prob = 0.5) # getting tails
barplot(fy, names = y, col = 'red', xlab = 'Y', ylab = 'probability', main = 'binomial, p = 0.5, 10 trials')

# increase prob
y = seq(0,10)
fy = dbinom(y, 10, prob = 0.9) # 90% of getting tails
barplot(fy, names = y, col = 'red', xlab = 'Y', ylab = 'probability', main = 'binomial, p = 0.9, 10 trials')

# lower prob
# increase prob
y = seq(0,10)
fy = dbinom(y, 10, prob = 0.1)
barplot(fy, names = y, col = 'red', xlab = 'Y', ylab = 'probability', main = 'binomial, p = 0.1, 10 trials')
```

Poisson distribution
- number of individuals, arrivals, events, coutns, etc. in a given time/space/unit of counting effort, if each event is independent of all others
- discrete
- range: non-negative integers
- parameters: lambda (rate) [mean and variance]
```{r}
y = seq(0,10)
fy = dpois(y, lambda = 1)
barplot(fy, names = y, col = 'red', xlab = 'Y',
        ylab = 'probability', main = 'poisson, lambda = 1')

# lambda = 0.1
y = seq(0,10)
fy = dpois(y, lambda = 0.1)
barplot(fy, names = y, col = 'red', xlab = 'Y',
        ylab = 'probability', main = 'poisson, lambda = 0.1')

```

## how to interpret your console

> it's all good, you're code ran!
+ something is missing from your code, could be unmatched bracket, or bad internet connection (so just restart your R session to fix that problem)