---
title: "Week 3 in class lectures"
author: "sbsambado"
date: "4/13/2020"
output: html_document
---

Topic 2: Normal distribution

Old final exam questions

6. The weights of all females fruit flies in a very large population of fruit flies are normally distributed with a mean of 0.3 mg and a standard deviation of 0.01 mg. We select a female fruit fly at random from the population. What is the probability that it has a weight of less than 0.27 mg or more than 0.31 mg?
```{r}
# plotting the probability density function for a standard normal distribution

mu = 0.3 # set mean
sd = 0.01 # set standard deviation

flyweight = seq(mu - (5*sd), mu + (5*sd), 0.001) # creates a sequence of numbers from mu-5*sd to mu+5*sd, at intervals of 0.001

prob = dnorm(flyweight, mean=mu, sd=sd) # caclulates the probability density for every value of Y

# make a plot with X on the x-axis and fx on the y-axis
plot(flyweight, prob, type = 'l', ylab = 'probability density of weight', xlab = 'weight (mg)', main = 'fruitflies')
abline(v = 0.27, col = 'red')
abline(v = 0.31, col = 'blue')

pnorm(0.27, mean = 0.3, sd = 0.01) +
  (1 - pnorm(0.31, mean = 0.3, sd = 0.01))
```

Topic 3: Inference about mean

95% confidence interval for a mean

Paradise flying snakes
```{r}
sample = c(0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6)
estimate_mean = 1.375
sd = 0.324
n = 8

x = seq(-5, 5, 0.1)
fx = dt(x, df = 7) # df = degrees of freedom
plot(x, fx, xlab = 'X', ylab = 'probability density', main = 't distribution, df = 7',
     type = 'l')
abline(v = qt(0.975, df = 7), col = 'red') # quantile function

qt(0.975, df = 7)

lower = 1.375 - qt(0.975, df = 7)*0.324/sqrt(8)
lower

upper = 1.375 + qt(0.975, df = 7)*0.324/sqrt(8)
upper


## easier way to do this, One Sample t-test
Y = c(0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6)
t.test(Y)
```

Topic 5: One sample t-test

> dataframe <- c(data points)
> mu0 = mean from sampled data frame

>t.test(dataframe, mu = mu0)


Assumptions:
1. variable is normally distributed
2. sample is a random sample

Human body temperature
```{r}
# find the p-values

mu0 = 98.6
n = 24
Ybar = 98.28
s = 0.94

se = s/sqrt(n)
df = n - 1

t = (Ybar - mu0)/se

# draw the t distribution
curve(dt(x, df = 23), xlim = c(-5,5), xlab = 't', ylab = 'probability density',
      main = 't-distribution', df = 23)
abline(h = 0, col = 'grey')

# define left shaded region
xmin <- -5
lowerbound <- t

xpoints <- c(xmin, seq(xmin, lowerbound, 0.01), lowerbound)
ypoints <- c(0, dt(seq(xmin, lowerbound, 0.01), df = 23), 0)
polygon(xpoints, ypoints, col = 'red')

# define right shaded region
xmax <- 5
upperbound <- -t

xpoints2 <- c(upperbound, seq(upperbound, xmax, 0.01), xmax)
ypoints2 <- c(0, dt(seq(upperbound, xmax, 0.01), df = 23), 0)
polygon(xpoints2, ypoints2, col = 'red')

abline(v = qt(0.025, df = 23), lty = 2)
abline(v = qt(0.975, df = 23), lty = 2)

pvalue = 2*pt(t, df = 23)
pvalue


## but increase your sample size to improve your pvalue
# find the p-values

mu0 = 98.6
n = 130
Ybar = 98.28
s = 0.94

se = s/sqrt(n)
df = n - 1

t = (Ybar - mu0)/se

# draw the t distribution
curve(dt(x, df = 129), xlim = c(-5,5), xlab = 't', ylab = 'probability density',
      main = 't-distribution', df = 129)
abline(h = 0, col = 'grey')

# define left shaded region
xmin <- -5
lowerbound <- t

xpoints <- c(xmin, seq(xmin, lowerbound, 0.01), lowerbound)
ypoints <- c(0, dt(seq(xmin, lowerbound, 0.01), df = 129), 0)
polygon(xpoints, ypoints, col = 'red')

# define right shaded region
xmax <- 5
upperbound <- -t

xpoints2 <- c(upperbound, seq(upperbound, xmax, 0.01), xmax)
ypoints2 <- c(0, dt(seq(upperbound, xmax, 0.01), df = 129), 0)
polygon(xpoints2, ypoints2, col = 'red')

abline(v = qt(0.025, df = 129), lty = 2)
abline(v = qt(0.975, df = 129), lty = 2)

pvalue = 2*pt(t, df = 129)
pvalue
```

