---
title: "Week 3 - Lecture material"
author: "sbsambado"
date: "4/15/2020"
output: html_document
---
Following along zoom meeting with Dr. Briggs for Biometry

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

April 15th (Wednesday)

## Goals
1. To illustrate some of the things you can do in R with the Normal and t distributions.

### Normal Distribution in R
`pnorm` (`p` for "probability"): cumulative distribution function, cdf
`qnorm`(`q` for "quantile"): inverse cdf
`dnorm` (`d` for "density"): the density function, pdf
`rnorm` (`r` for "random"): a random number from that distribution

## Week 3, topic 2: British Spy example
MI5 says a man has to be shorter than 180.3 cm tall to be a spy. (5’ 11”)

Mean height of British men is 177.0cm, with standard deviation 7.1cm, with a normal distribution.

What proportion of British men are excluded from a career as a spy by this height criteria?

```{r}
#Plotting the probability density function for a Standard Normal Distribution. 
mu = 177  # set the mean
sd = 7.1  # set the standard deviation

height = seq(mu-(5*sd), mu+(5*sd), 0.1)  # This creates a sequence of numbers from mu-5*sd to mu+5*sd, at intervals of 0.1

prob = dnorm(height, mean=mu, sd=sd)  # this calculates the probability density for every value of Y

# Make a plot with X on the x-axis and fx on the y-axis.
# the option type = "l" says to plot it as a line graph (the default is points)  
plot(height, prob, type = "l", ylab="probability density of height", xlab="height (cm)", main="Height of British men")
abline(v=180.3,col="red")

```

Calculate the probability of height > 180.3 cm
```{r}

1 - pnorm(180.3,mean = mu, sd = sd)

```

Remember that the `pnorm` command is the cumulative probability density. It gives the probability of being less than or equal to a given value:

```{r}
cdf_height =  pnorm(height, mean=mu, sd=sd)

plot(height,cdf_height, type = "l", ylab="cdf(height)",
 xlab="height", main="Normal distribution")
abline(v=180.3, col="red")
abline(h=pnorm(180.3,mean = mu, sd = sd), col="blue")
```


## NASA Astronaut Example
To be a NASA astronaut, you must be between 148.6 cm and 193.0 cm in height.
What fraction of American females are eligible to be an astronaut, if the mean height of female Americans is 162.6 cm and the standard deviation is 6.4 cm?

```{r}
#Plotting the probability density function for a Standard Normal Distribution. 
mu = 162.6  # set the mean
sd = 6.4  # set the standard deviation

height = seq(mu-(5*sd), mu+(5*sd), 0.1)  # This creates a sequence of numbers from mu-5*sd to mu+5*sd, at intervals of 0.1

prob = dnorm(height, mean=mu, sd=sd)  # this calculates the probability density for every value of Y

# Make a plot with X on the x-axis and fx on the y-axis.
# the option type = "l" says to plot it as a line graph (the default is points)  
plot(height, prob, type = "l", ylab="probability density of height", xlab="height (cm)", main="Height of American women")
abline(v=148.6,col="purple")
abline(v=193.0,col="orange")
```



```{r}
cdf_height =  pnorm(height, mean=mu, sd=sd)

plot(height,cdf_height, type = "l", ylab="cdf(height)",
 xlab="height", main="Normal distribution")
abline(v=148.6,col="purple")
abline(v=193.0,col="orange")
```

```{r}
pnorm(193, mean = 162.6, sd = 6.4) - pnorm(148.6, mean = 162.6, sd = 6.4)
```

## Fruit fly example. 
The weights of all female fruit flies in a very large population of fruit flies are normally distributed with a mean of 0.3 mg and a standard deviation of 0.01 mg. We select a female fruit fly at random from the population.  What is the probability that it has a weight of less than 0.27 mg or more than 0.31 mg? 

```{r}
#Plotting the probability density function for a Standard Normal Distribution. 
mu = 0.3  # set the mean
sd = 0.01  # set the standard deviation

flyweight = seq(mu-(5*sd), mu+(5*sd), 0.001)  # This creates a sequence of numbers from mu-5*sd to mu+5*sd, at intervals of 0.001

prob = dnorm(flyweight, mean=mu, sd=sd)  # this calculates the probability density for every value of Y

# Make a plot with X on the x-axis and fx on the y-axis.
# the option type = "l" says to plot it as a line graph (the default is points)  
plot(flyweight, prob, type = "l", ylab="probability density of weight", xlab="weight (mg)", main="fruit flies")
abline(v=0.27,col="red")
abline(v=0.31,col="blue")
```

Calculate the probability of being in either of these two tails: 
```{r}
pnorm(0.27,mean=0.3,sd=0.01) + (1- pnorm(0.31,mean=0.3,sd=0.01) )
```

Or, you can use the "lower.tail = FALSE" option to calculate the probability of being above a given value.
```{r}
pnorm(0.27,mean=0.3,sd=0.01) + pnorm(0.31,mean=0.3,sd=0.01,lower.tail = FALSE)
```

## Week 3, topic 3: Inference about the mean
## Westie Example
The distribution of weights of all West Highland Terriers in the world is well approximated by a normal distribution with a mean of 18 pounds, and a standard deviation of 2 pounds.  

West Highland Terriers that are heavier than 21 pounds are considered overweight. What fraction of West Highland Terriers are considered overweight? 

This is just another example like the ones above. 


```{r}
#Plotting the probability density function for a Standard Normal Distribution. 
mu = 18.0  # set the mean
sd = 2.0  # set the standard deviation

# whole population 

weight = seq(mu-(5*sd), mu+(5*sd), 0.1)  # This creates a sequence of numbers from mu-5*sd to mu+5*sd, at intervals of 0.1

prob = dnorm(weight, mean=mu, sd=sd)  # this calculates the probability density for every value of Y

# Make a plot with X on the x-axis and fx on the y-axis.
# the option type = "l" says to plot it as a line graph (the default is points)  
plot(weight, prob, type = "l", ylab="probability density of weighteight", xlab="weight (pounds)", main="Weight of Westies")
abline(v=21.0,col="green")
```
Calculate the probability of a weight greater than or equal to 21:
```{r}
1-pnorm(21, mean = 18, sd = 2)
```
or:
```{r}
pnorm(21, mean = 18, sd = 2, lower.tail = FALSE) # lower.tail FALSE means we want to know whats that value and above, can omit 1 -
```


The next part of the question, rather than asking about the weight of a random individual in the population (or about the fraction of the population), asks about the *mean of a sample of a given size*. 

The distribution of weights of all West Highland Terriers in the world is well approximated by a normal distribution with a mean of 18 pounds, and a standard deviation of 2 pounds.  

We record the weight of 9 randomly sampled West Highland Terriers. What is the probability that the *mean* weight of our sampled West Highland Terriers is less than 17 pounds? 

The distribution of *mean* weights takes on a narrower distribution than the distribution of random values in the population. 

This is the sampling distribution for the mean of a sample of 9 westies.

```{r}
#Plotting the probability density function for a Standard Normal Distribution. 
mu = 18.0  # set the mean
sd = 2.0  # set the standard deviation
n = 9 # the more you increase this, the more narrow the peak
# distribution of means, gives a more narror, if we sampled multiple times, we would get the mean
se = sd/sqrt(n)

weight = seq(mu-(5*sd), mu+(5*sd), 0.1)  

prob = dnorm(weight, mean=mu, sd=se) # sd is now se

# Make a plot with X on the x-axis and fx on the y-axis.
# the option type = "l" says to plot it as a line graph (the default is points)  
plot(weight, prob, type = "l", ylab="probability density of weighteight", xlab="weight (pounds)", main="Mean weight of n = 9 Westies")
abline(v=17.0,col="brown")
```
We can calculate the probability that the mean weight of a sample size of 9 that is less than 17:
```{r}
pnorm(17,mean = 18, sd = 2/sqrt(9)) # sd = se = 2/sqrt(9)
```

### Student's t Distribution in R
`pt` (`p` for "probability"): cumulative distribution function, cdf
`qt`(`q` for "quantile"): inverse cdf
`dt` (`d` for "density"): the density function, pdf
`rt` (`r` for "random"): a random number from that distribution

Let's compare the probability density function (pdf) of a Student's t distribution with different degrees of freedom, with the pdf of a Standard Normal Distribution:
```{r}
mu = 0  # set the mean
sd = 1  # set the standard deviation

x = seq(mu-(5*sd), mu+(5*sd), 0.1) 
plot(x, dnorm(x,mean=mu,sd = sd), type = "l", ylab="probability density", xlab="x", main="")
lines(x,dt(x,df=30),col="green")
lines(x,dt(x,df=5),col="red")
lines(x,dt(x,df=2),col="blue")

```

## Calculating Confidence Intervals


# Quantiles, `qt`
Quantiles are the inverse of the cumulative distribution function.  
Whereas the cumulative distribution function tells us the probability that X is less than or equal to a specified value, quantiles tell us the value of X for which there is a given probability of getting a value that is less than or equal to that value.  For example, the 50% quantile is the value of X for which 50% of the probability is below that value, and 50% is above that value.
The 10% quantile is the value of X for which 10% of the probability is below that value. 



# The Paradise flying snake example
Ungulation rates of snakes: 
The sample:  0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6
Calculate the 95% Confidence interval for the mean ungulation rate

```{r}
x = seq(from=-5,to=5,by=0.1)
fx = dt(x,df=7)
plot(x,fx,type="l",col="black",xlab = "t", ylab = "probability density", main = "t distribution, df=7")
abline(v=qt(0.975,df=7),col="red")
abline(v=qt(0.025,df=7),col="blue")
```
We can use Quantiles, `qt` to figure out the value of t at which 2.5% of the probability is above the red line, and 2.5% of the probability is below the blue line.  

```{r}
# 97.5% quantile for a t-distribution with 7 degrees of freedom

qt(0.975,df=7)

```


```{r}
# 2.5% quantile for a t-distribution with 7 degrees of freedom
qt(0.025,df = 7)
```



calculate the lower bound of the 95% confidence interval 
```{r}
ungulation = c(0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6)
n = length(ungulation) # length command counts number of values
df = n - 1
mean = mean(ungulation)
se = sd(ungulation)/sqrt(n)
lower = mean - qt(0.975,df=df)*se
lower
```
or alternatively:
```{r}
lower = mean + qt(0.025,df=df)*se
lower
```

calculate the upper bound of the 95% confidence interval:
```{r}
upper = mean + qt(0.975,df=df)*se
upper

```

the t-test will give this to you automatically
```{r}
ungulation = c(0.9, 1.4, 1.2, 1.2, 1.3, 2.0, 1.4, 1.6)
t.test(ungulation)
```

99% confidence interval:
```{r}
t.test(ungulation, conf.level = 0.99)
```