---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

Read in frog data
```{r}

frogs <- read.csv("~/Desktop/allpit_5yr.csv")
head(frogs)

# Julian data is treated as continual variable from some date
```


```{r}
Unicorn<-frogs[frogs$Site_name=="UnicornBasin",]
Mono<-frogs[frogs$Site_name=="MonoPass",]
LIV<-frogs[frogs$Site_name=="LittleIndianValley",]

```

Pretend we want to do a simple analysis comparing the Bd loads between these 3 sites, and look for an effect of snout-to-vent length (SVL) on Bd load.  But, General linear models assume that the error (unexplained variance) is normally distributed.
Let's look at the Bd loads at these 3 sites:
```{r}
hist(Unicorn$Zoospores,50)
```

```{r}
mean(Unicorn$Zoospores)
median(Unicorn$Zoospores)

min(Unicorn$Zoospores)
max(Unicorn$Zoospores)
length(Unicorn$Zoospores)
length(Unicorn$Zoospores[Unicorn$Zoospores==0])
length(Unicorn$Zoospores[Unicorn$Zoospores==0])/length(Unicorn$Zoospores)

```


```{r}
hist(Mono$Zoospores,50)
```

```{r}
min(Mono$Zoospores)
max(Mono$Zoospores)
length(Mono$Zoospores)
length(Mono$Zoospores[Mono$Zoospores==0])
length(Mono$Zoospores[Mono$Zoospores==0])/length(Mono$Zoospores)

# highly skewed data
```

```{r}
hist(LIV$Zoospores,50)
```

```{r}
mean(LIV$Zoospores)
median(LIV$Zoospores)

min(LIV$Zoospores)
max(LIV$Zoospores)
length(LIV$Zoospores)
length(LIV$Zoospores[LIV$Zoospores==0])
length(LIV$Zoospores[LIV$Zoospores==0])/length(LIV$Zoospores)

# 38% swabs are 0
```

All of these data are clearly not Normally distributed.
With so many zeros, we can't just log transform and add a small number.
This just creates a peak at the log of whatever small number we add:
```{r}
# log transformed data still gives huge peak
smallnumber=0.01 # adding 1 adds a lot of fraction to the dataset since this dataset has small numbers
hist(log(Unicorn$Zoospores+smallnumber),50)
```

# Quick and dirty approach
We could try dividing our analysis into 2 parts, and comparing the fraction infected between the 3 populations, and then comparing the Bd load between the 3 populations.
*What cut-off do we use to declare an individual infected?* 
Zero Zoospores? 
1 Zoospore? 

We'll create 2 new variables in frogs: 

Infected0 is 1 if Zoospores > 0 (especially if you get multiple 0s)

Infected1 is 1 if Zoospores > 1

```{r}
# added column to dataset
frogs$Infected0<-ifelse(frogs$Zoospores>0,1,0)
frogs$Infected1<-ifelse(frogs$Zoospores>1,1,0)
```

Logistic regression to compare Bd prevalence at the 3 sites:
First for Infected0

```{r}
#explain infected as a function of site name + snout length
# family comes from binomial distribution 1/0
prevmod0 <- glm(Infected0 ~ Site_name+SVL, data = frogs, family = "binomial")
summary(prevmod0)
# SNVL no affect whether infected or not
# different prevalences at different sites
# logit transformed, interprest the estimate as pop infected
# intercept is site little indian valley (alphabetica)
```
Logistic regression to compare Bd prevalence at the 3 sites:
Then for Infected1

```{r}
# cut off of 1
# finds differences between prevalences of different sites, gives slightly different results
prevmod1 <- glm(Infected1 ~ Site_name+SVL, data = frogs, family = "binomial")
summary(prevmod1)
```

frogload0 is a dataset with only Bd-positive frogs, using the 0 zoospore cut-off.

frogload1 is a dataset with only Bd-positive frogs, using the 1 zoospore cut-off.



```{r}
frogload0<-frogs[frogs$Zoospores>0,]
frogload1<-frogs[frogs$Zoospores>1,]
```

Create new variable that is log of zoospores in each of these datasets:

```{r}
# dont have to add values since they've all be transformed to be greater than 0
frogload0$logZoospores = log(frogload0$Zoospores)
frogload1$logZoospores = log(frogload1$Zoospores)
```

Use lm to compare Bd loads between sites for only the infected frogs.
First using the 0 zoospore cut-off
```{r}
loadmod0<-lm(logZoospores ~ Site_name+SVL, data = frogload0)
summary(loadmod0)
# here shows effect of SVL
```

Then using the 1 Zoospore cut off:
```{r}
# cut off 1
# SVL effect goes away
loadmod1<-lm(logZoospores ~ Site_name+SVL, data = frogload1)
summary(loadmod1)
```


Let's try the Zero-inflated models.  The models that are based on Poisson and Negative Binomial distributions are expecting count data.  That is, they are expecting integer values only. 
We still need to decide what to do for the values of zoospores between 0 and 1.  
Are these individuals infected or uninfected?  Should we include these individuals in the 0 category, or in the 1 category????

Let's try both options. 

```{r}
frogs$Zoosp_round_down <- floor(frogs$Zoospores) # round down
frogs$Zoosp_round_up <- ceiling(frogs$Zoospores) # round up
```

## ZIP and ZINB models
Hurdle model : treat as 2 part process. 1 part is 0 vs non-xeros. 2nd part is
```{r}
#install.packages("pscl")
#install.packages("lmtest")
library(pscl)
library(lmtest)

```
Zero inflated Poisson:

2 type of 0s: good and bad zeroes 
```{r}
# posi: variance = mean 

f1down <- formula(Zoosp_round_down ~ Site_name + SVL | Site_name + SVL) # what explains the load that's not 0 # | explantory variables if something is infected or not
ZIP_round_down <- zeroinfl(f1down, dist = "poisson", link = "logit", data = frogs)
summary(ZIP_round_down)

# make simpler to make model run 
f1down <- formula(Zoosp_round_down ~ Site_name + SVL | Site_name) # what explains the load that's not 0 # | explantory variables if something is infected or not
ZIP_round_down <- zeroinfl(f1down, dist = "poisson", link = "logit", data = frogs)
summary(ZIP_round_down)
```

Zero inflated Negative Binomial:

```{r}
# negative binom: variance >> mean
f1down <- formula(Zoosp_round_down ~ Site_name + SVL | Site_name + SVL)
ZINB_round_down <- zeroinfl(f1down, dist = "negbin", link = "logit", data = frogs) # changed distribution
summary(ZINB_round_down)
# doesnt fit
# this is the first way to use bayesian approaches to fit things, nt enough combos of infected + SVL

# but if you make it a simpler model it will run
f1down <- formula(Zoosp_round_down ~ Site_name + SVL | Site_name )
ZINB_round_down <- zeroinfl(f1down, dist = "negbin", link = "logit", data = frogs) # changed distribution
summary(ZINB_round_down)
```
Likelihood ratio test to compare ZIP and ZINB
```{r}
# test the two models
# sys neg binomial is better model for this data
lrtest(ZIP_round_down,ZINB_round_down)
```
Overwhelming evidence to go with ZINB, rather than ZIP


# ZIP and ZINB on zoospores_round_up

Rounding data up

Zero inflated Poisson:

```{r}
f1up <- formula(Zoosp_round_up ~ Site_name + SVL | Site_name)
ZIP_round_up <- zeroinfl(f1up, dist = "poisson", link = "logit", data = frogs)
summary(ZIP_round_up)
# has toruble fitting
```

Zero inflated Negative Binomial:

```{r}
f1up <- formula(Zoosp_round_up ~ Site_name + SVL | Site_name)
ZINB_round_up <- zeroinfl(f1up, dist = "negbin", link = "logit", data = frogs)
summary(ZINB_round_up)
```


Likelihood ratio test to compare ZIP and ZINB
```{r}
lrtest(ZIP_round_down,ZINB_round_down)
```
Overwhelming evidence to go with ZINB, rather than ZIP



###Zero Adjusted Poisson and Negative Binomial (hurdle models)
Now try hurdle models

# ZAP and ZANB on Zoosp_round_down
```{r}
H1downA <- hurdle(f1down, dist = "poisson", link = "logit",data = frogs)
H1downB <- hurdle(f1down, dist = "negbin", link = "logit",data = frogs)
lrtest(H1downA,H1downB)

# neg bionomial better
```
Definitely need to use Negative Binomial version

```{r}
summary(H1downB)
```

# ZAP and ZANB on Zoosp_round_up
```{r}
H1upA <- hurdle(f1up, dist = "poisson", link = "logit",data = frogs)
H1upB <- hurdle(f1up, dist = "negbin", link = "logit",data = frogs)
lrtest(H1upA,H1upB)
```

Definitely need to use Negative Binomial version

```{r}
summary(H1upB)
```
